---
title: "Air Dataset"
author: "Lakshita"
date: "2025-11-10"
output:
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
  pdf_document:
    toc: true
    number_sections: true
    toc_depth: 3
---


# Loading Libraries
```{r warning=FALSE,message=FALSE}
library(ggplot2)
library(maps)
library(ggrepel)
library(tidyverse)
library(dplyr)
library(tidyr)
library(corrplot)
library(gridExtra)
library(plotly)
library(factoextra)
library(psych)
library(GGally)
```

# Loading data
```{r}
air=read.csv('indian_weather_data.csv')
head(air)

# making numerical df
num_df=air[c("lat","lon","temperature","weather_code","co","no2","o3","so2","pm2_5","pm10","wind_speed","wind_degree","pressure",    "precip","humidity","cloudcover","feelslike","visibility" )]
num_df=data.frame(num_df)

# making categorical df
cat_df=air[c("city","sunrise","sunset","moonrise","moonset","wind_dir")]
```
# Data Handling
```{r}
colSums(is.na(air))
```
# PCA: Principal Component Analysis
```{r}
pca_fit <- prcomp(num_df, scale=TRUE)
pca_summary<-summary(pca_fit)

importance_matrix<-pca_summary$importance

# Convert to data frame
pca_df <- as.data.frame(importance_matrix)
pca_df

# Rotating the factors
rotation_df<-as.data.frame(pca_fit$rotation[,0:5])
rotation_df
```

```{r}
fviz_pca_biplot(pca_fit, 
                repel = TRUE,  # Prevents text overlapping
                col.var = "blue",    # Variables color
                col.ind = "gray",    # Individuals color
                title = "PCA Biplot")
```

# Factor Analysis

## KMO and Bartlett's Test of Sphericity 
```{r}
# making a correlation matrix
correlation_matrix<-cor(num_df)

#KMO Test
kmo_result <- KMO(correlation_matrix)
print(kmo_result)

# Bartlett's Test of Sphericity 
cortest.bartlett(correlation_matrix, n = nrow(num_df))
```
1. **KMO Test :** we can observe the overall MSA value is greater than 0.69 indicating that correlation matrix is not identity matrix.
2. **Bartlett's Test of Sphericity :** We can Observe that the p-value is 1.449223e-293 <0.05 indicating that the data is suitable for factor analysis

## Deciding number of factors
```{r}
scree(num_df, factors = FALSE, pc = TRUE, 
      main = "Scree Plot for Factor Analysis")
```


```{r}
fa<- fa(num_df, nfactors=5,rotate="varimax",scores="regression")
fa
```


Factors are:

1. Factor 1: positive effect: co, o3, so2, pm2_5,pm10
             negative effect: lon, temprature
             
2. Factor 2: positive effect: lat,lon,pressure
             negative effect: temperature,feelslike
             
3. Factor 3: positive effect: wether_code,humidity
             negative effect: visibility

4. Factor 4: positive effect: no2,wind_speed

5. Factor 5: positive effect: precip

* Factor Analysis is used for identifying the **underlying structure of the data**. It helps in reducing variable and these obtained factors can be effectively used for EDA.

* The First Factor can be named as pollutants, Second Factor as geographic_cond, Third Factor is weather_cond , Fourth can be named as ozone since higher wind speed decreases no2 concentration and fifth can be named precipitation

# Data Engineering

## Adding FA Score / Factors for EDA
```{r}
# Storing FA Scores as df 
fa_scores<-as.data.frame(fa$scores)

# renaming column names
colnames(fa_scores)=c("pollutants","geographic_cond","weather_cond","ozone","precipitation")

# concatenating 2 dfs air and fa_scores
df<-cbind(air,fa_scores)
head(df)
```
## Converting Weather code and visibility to categorical variables
```{r}
df<-df %>% 
  mutate(weather_cat=case_when(
    weather_code==113~"sunny",
    weather_code==122~"partly cloudly",
    weather_code==143~"mist",
    weather_code==116~"moderate rain",
    weather_code==119~"showers",
    weather_code==248~"fog",
    weather_code==176~"moderate rain",
  )) %>% 
  mutate(visibility_cat=case_when(
    visibility<1~"very poor",
    between(visibility,1,3) ~"poor",
    between(visibility,3,5) ~"moderate",
    visibility>5 ~"good"
  )) %>% 
  mutate(parts_of_India = case_when(
    between(lat, 28, 37.6) & between(lon, 68.7, 97.25) ~ "North",
    between(lat, 15, 28) & between(lon, 68, 78) ~ "West",
    between(lat, 20, 28) & between(lon, 83, 97.25) ~ "East",
    lat < 20 & between(lon, 74, 84) ~ "South",
    between(lat, 18, 26) & between(lon, 74, 85) ~ "Central",
    between(lat, 22, 28) & between(lon, 89, 97.25) ~ "Northeast",
    TRUE ~ "Other Region"
  ))
  
```


# Exploratory Data Analysis

## Correlation among variables 
```{r}
# Calculate correlation with pairwise complete observations
weather_cor <- cor(num_df,
                   use = "pairwise.complete.obs")

c_color<- colorRampPalette(c("hotpink", "lightgreen","darkblue"))  

corrplot(weather_cor, 
         method = "color",
         title = "Weather Variables Correlation",
         order="hclust",
         col=c_color(10),
         tl.col="black"
         )
```
* Positive Correlation: pollutants are highly correlated such as pm2_5, pm_10, co, so2 , no2 is moderately correlated, feelslike and temperature

* Moderate Correlation (positive and negative): We can see moderate correlation between co2 and weather code,longitude with temperature and feelslike,latitude with temprature and feelslike

* Negative Correlation:pressure with feelslike and temprature 


## Temprature according to latitude and longitude values
```{r}
world<-map_data("world")

#getting the map for india 
india<-subset(world,region=="India")


ggplot()+
   geom_polygon(data=india,
                aes(x=long,y=lat,group=group))+
    geom_point(data=df,
             aes(x=lon,y=lat,color=pm2_5))+
  scale_color_continuous(
    type = "viridis",  # Or "gradient"
    name = "pm 2.5"
  )+
  scale_x_log10()
```

```{r}
# setting theme for all plots
set_theme(theme_minimal()+
            theme( 
                plot.title=
                  element_text(
                    size=rel(2)),
                
                panel.background = 
                  element_rect(color="black"),
                
              ))
```

## Top 20 cities with worst air quality pm10 and pm2
```{r}
# pm10
df %>% 
  arrange(desc(pm10)) %>% 
  select(city,pm10,wind_dir) %>% 
  slice_head(n=20) %>% 
  ggplot(
    aes(x=reorder(city,-pm10),y=pm10,fill=wind_dir)
  )+
  labs(
    title="Highest pm10 Vs Cities and their wind direction",
    x="Cities",
    y="PM 10"
  ) +
  geom_bar(stat="identity")+
  theme(
      axis.text.x = 
                  element_text(angle=45,
                                hjust=1,
                               face="bold")
  )


```
```{r}

df %>% 
  arrange(desc(pm2_5)) %>% 
  select(city,pm2_5,wind_dir) %>% 
  slice_head(n=20) %>% 
  ggplot(
    aes(x=reorder(city,-pm2_5),y=pm2_5,fill=wind_dir)
  )+
  geom_bar(stat="identity")+
  labs(
    title="Highest pm 2.5 Vs Cities and their wind direction",
    x="Cities",
    y="PM 2.5"
  )+
  theme(
      axis.text.x = 
                  element_text(angle=45,
                                hjust=1,
                               face="bold")
  )
```
#One-on-one relationships between two continuous variables
## Temperature vs PM2.5 levels
```{r}

# temperature Vs pm2.5 levels
plot1<-df %>% 
  ggplot(aes(temperature,pm2_5,size=visibility))+
  geom_point(color="orange")+
  labs(
    title="Temperature Vs pm 2.5",
    subtitle="There influence on Visibility",
    x="Temperature",
    y="pm 2.5"
  )+
  theme(
    aspect.ratio = 1
  )
plot1

```

## Temptrature Vs Pressure 
```{r}

# temperature Vs pm2.5 levels
plot2<-df %>% 
  ggplot(aes(temperature,pressure,size=visibility))+
  geom_point(color="pink")+
  labs(
    title="Temperature Vs Pressure",
    subtitle="There influence on Visibility",
    x="Temperature",
    y="Pressure"
  )+
  theme(
    aspect.ratio = 1
  )

plot2
```
# Observing 3 variables Wind speed , Temprature and pm 2.5 concentration
```{r}
fig <- plot_ly(df, x = ~wind_speed, y = ~temperature, z = ~pm2_5, color = ~city)
fig
```
# Clustering based on pollutants

## Making dataset for clustering
```{r}
# Data for clustering
k_data<-df %>% 
  select("city","pollutants","weather_cond","geographic_cond")
```
## Selecting number of clusters
```{r}
fviz_nbclust(k_data[,c("pollutants","weather_cond","geographic_cond")], kmeans, method = "wss") + 
  ggtitle("Elbow Method")

fviz_nbclust(k_data[,c("pollutants","weather_cond","geographic_cond")], kmeans, method = "silhouette") + 
  ggtitle("Silhouette Method")
```
## Performing K-means clustering
```{r}
# Perform k-means clustering (e.g., 4 clusters)
set.seed(123)
kmeans_result <- kmeans(k_data[,c("pollutants","weather_cond","geographic_cond")], centers = 4, nstart = 25)
print(kmeans_result)

```

## Storing Clusters as factor
```{r}
k_data$cluster <- as.factor(kmeans_result$cluster)

k_data %>%
  mutate(cluster_name = case_when(
    cluster == 1 ~ "Extremely Clean, Very Warm",
    cluster== 2 ~ "Outlier City",
    cluster== 3 ~ "Clean, Bad Weather",
    cluster== 4 ~ "Very Polluted, Bad Weather",
    cluster== 5 ~ "Polluted, Good Weather",
    TRUE ~ "Unknown"
  ))

```
## Visualizing clusters
```{r}
city_names<-k_data$city
fviz_cluster(kmeans_result, data = k_data[,c("pollutants","weather_cond","geographic_cond")],
             palette = "Set2", ggtheme = theme_minimal(),
             geom = "point") +
  geom_text(aes(label = city_names), 
            check_overlap = TRUE, 
            size = 3, 
            vjust = -0.5)+
 labs(
   title="Clusters Visualization"
 )+
  theme(
    plot.title = 
      element_text(face = "bold",
                   size=rel(2)),
    panel.background = 
                  element_rect(color="black")
  )+
  scale_fill_discrete(labels = c("A", "B", "C","D","E"))
```
